Incorporating Periodic Variability in Hidden Markov Models for Animal Movement

### 2016-09-08 bmc_template/manuscript.tex
working on ms, deleted bmc_article

Given a sequence of animal movements, movement models aim to find a parsimonious description
that can be used to understand past movements and predict
future movements. Ecologists have long considered the effects of
individual-level covariates (sex, age, nutritional status) and
environmental covariates (habitat type, location of predators or prey)
on movement \cite{patterson2008state, mckenzie2009first, pal1998dispersal}. 
More recently, modelers have developed
\emph{hidden Markov models} (HMMs)
\cite{firle_influence_1998,nathan_movement_2008,langrock_flexible_2012} % 
--- used in animal ecology under the rubric of the ``multiphasic movement
framework'' \cite{fryxell_multiple_2008} --- that consider the effects
of organisms' \emph{internal} states; in particular, HMMs model animal
movement as though individual animals' movement behaviour 
at particular times is determined
by which of a discrete set of unobserved movement states
(e.g. ``foraging'', ``traveling'', ``resting'') they currently occupy.
Conditional on the state occupied by an individual, HMMs typically
assume that animals follow a standard correlated random walk model
\cite{okubo_diffusion_1980,turchin1998quantitative}.

Ever-increasing capabilities of remote
sensors are making movement data available over an
ever-wider range of time scales, at both higher resolution (e.g. hourly data
from GPS collars vs. daily or weekly fixes for radio or VHF
collars) and longer extent (e.g. from a few days to significant
fractions of a year, or longer).  When analyzing such
long-term data, ecologists will more
often have to account for temporal variability in movement
behaviour at diurnal and seasonal scales that were previously
not captured in the data.

@@

Our second goal, therefore, is to explore whether
van de Kerk et al.'s results on optimal
model complexity might be driven at
least in part by structural problems with their statistical
model, i.e. the
assumption of temporally homogeneous behaviour.  For large data sets, 
information-theoretic model selection methods will
typically choose complex, highly parameterized models; when there is
only one way in which models can become more complex (e.g. by
increasing the number of latent states), complexity that is
present in the data but not accounted for in the model (e.g. spatial
or temporal heterogeneity) can be misidentified as other forms of
complexity.  We predict that increasing volumes of data will
increasingly lead researchers who are accustomed to fitting 
small models to sparse data into such traps.  We examine whether
allowing for diurnal variation in the Florida panther data allows us to
select models with fewer latent states; we also
fit models to simulated data with varying numbers of latent states,
and with and without temporal heterogeneity, to test our conjecture that
heterogeneity can be misidentified as behavioural complexity.

@@

van de Kerk et al. \cite{kerk2015hidden} considered time-homogeneous models with 
a variety of candidate distributions %
--- log-Normal, Gamma, and Weibull distributions for step lengths
and von Mises and wrapped Cauchy distributions for the turning angle --- %
concluding on the basis of the Akaike information criterion (AIC) 
that Weibull step length and wrapped Cauchy turning angle distributions
were best.  
Since our analysis aims for simplicity and qualitative
conclusions rather than for picking the very best
predictive model, we focus on models that treat each step as
a univariate, log-Normally distributed observation, glossing
over both the differences in shape between
the three candidate step-length distributions and the effects
of considering multivariate (i.e., step length plus turning angle)
observations.  However, we do briefly compare log-Normal and Weibull
step-length distributions, with and without 
a von Mises-distributed turning angle included in the model (Figure 2).
(Note that most movement analyses, including van de Kerk et al.
\cite{kerk2015hidden}, are only partially multivariate, 
treating step length and turning angle at a particular time
as multivariate observations for the purpose of HMM
analysis but neglecting possible correlations
between the two measures.)

@@

Using BIC also simplifies 
evaluation of model selection procedures; 
it is easier to test whether our model selection
procedure has selected the model used to simulate the data,
rather than testing whether it has selected the model with
the minimal Kullback-Leibler distance
\cite{Richards2005}. We recognize that 
ecologists will often be interested in maximizing predictive
accuracy rather than selecting a true model, and that as
usual in ecological systems the true model will be far more
complicated than any candidate model \cite{BurnhamAnderson1998};
we believe that the qualitative conclusions stated here
for BIC-optimality will carry over to analyses using AIC instead.

@@

A simple simulation was conducted to to see whether fitting homogeneous transition probability HMMs to heterogeneous transitions simulated data will over estimate the number of states and estimate the correct number of states using the heterogeneous transitioning model. We used 100 realizations of a two-state HMM with sinusoidal temporal transitions and fitted it with 2- to 5-state HMMs with and without temporal heterogeneity in the transition probabilities.

@@

Model complexity and the number of parameters increase as the number of latent states increase.
For a fixed number of states homogeneous FMMs are simplest, followed by homogeneous HMMs and finally by FMMs and HMMs incorporating temporal heterogeneity. In general, the number of free parameters in an HMM is the sum of the number of free parameters for each of the three model components. Let $n$ be the number of hidden states and $k_{i}, k_{t}, k_{e}$ be the number of parameters describing the covariate-dependence of the prior distribution, transition function and emission distributions; that is, for a homogeneous model, $k=1$, while a single numeric covariate or a categorical predictor with two levels would give $k=2$. Then the number of free parameters of an HMM is:

\begin{equation}
\textrm{Number of Free Parameters}=\underbrace{k_{i}\cdot (n-1)}_{\text{Initial}} 
{+} \underbrace{k_{t}\cdot n\cdot (n-1)}_{\text{Transition}}
{+} \underbrace{ k_{e}\cdot n}_{\text{Emission}} .
\label{eq:numparam}
\end{equation}

As the number of states increases, the number of free parameters in (homogeneous or heterogeneous) FMMs and time-homogeneous HMMs will increase linearly, whereas for HMMs with temporal heterogeneity (or covariate-dependent transitions more generally) the number increases quadratically (Eq.~\ref{eq:numparam}). When comparing BICs, it is important to account for the tradeoff between log-likelihood and number of states, but also log-likelihood and number of free parameters.

We used simulations to predict expected hourly step lengths and
autocorrelation functions (ACF).  While the computation of expected
step length and ACF is straightforward for FMMs, and feasible for
homogeneous HMMs, the interaction between the geometric dwell time
within each state and the temporally varying interaction probabilities
makes it infeasible for more complex models.  We used this approach to
validate our models, comparing our simulated predictions with the
observed movements.  The more usual approach, generating predictions
from the expected step lengths conditional on the most likely state
sequence predicted by the Viterbi algorithm or pseduo-residuals \cite{zucchini_hidden_2009,langrock_flexible_2012}, is somewhat problematic
because the predictions by these methods rely on the observed data. These approach is useful to predict
missing data in the observation sequence, but because it is
conditional on the observed values, it can not reliably evaluate
goodness of fit for the different structural complexities of HMM
models.

@@

%% fig 1 goes somewhere here
The simulation agrees with our hypothesis that homogeneous transition HMMs can over estimate the number of hidden states. Heterogeneous transition models can always predict the correct number of states, whereas the temporally homogeneous models overestimate the number of states (based on BIC-optimality: Figure 1).

@@

Finally, if we want to estimate the number of states, BIC is not necessarily 
good for estimation of number of states \cite{biernacki2000assessing}, 
but it can be useful as an approximate upper limit estimate. This is a common problem when using BICs, thus, the "knee point" (jump point) was proposed for BIC in partitioning based clustering via angle-based method \cite{zhao_knee_2008}. A similar method was used with likelihood \cite{dean2012behavioural} where both are applicable in our case, and suggest fewer number of states, but we do not know the reliability when fitting more complex models (ie. the shape of the BIC curve can be affected by the non-linear increase of parameters with every additional state).

Incorporating temporal heterogeneity in animal movement is one step in
the right direction, but much remains to be done. 
Our model neglects other predictors, such as habitat type or
location with respect to environmental features such as roads,
that can potentially improve goodness of fit and predictions and 
further reduce the estimated number of states. While adding
more covariates is in principle straightforward using existing
frameworks, including all possible biological complexities in 
a HMM with state-dependent transitions may rapidly become intractable
in terms of both computational time and complexity of choosing
among possible reduced models and numbers of states.
Better diagnostic procedures and tests are needed:
these can both test overall goodness-of-fit \cite{potts_generalized_2014} 
and, more importantly, localize fitting problems to particular aspects of the
data so that models can be constructed without needing to
include all possible features of interest. Because there are a huge number of potential complexities that can be added to movement models (e.g. spatial/temporal/among-individual heterogeneity; effects of conspecific attraction or avoidance; memory or cognitive effects), each with associated costs in researcher and computational effort, such diagnostic plots are invaluable.

@@

>>>>

Given a sequence of animal movements, movement models aim to find a parsimonious description
that can be used to understand past movements and predict
future movements. Ecologists have long considered the effects of
individual-level covariates (sex, age, nutritional status) and
environmental covariates (habitat type, location of predators or prey)
on movement \cite{patterson2008state, mckenzie2009first, pal1998dispersal}. 
More recently, modelers have developed
\emph{hidden Markov models} (HMMs)
\cite{firle_influence_1998,nathan_movement_2008,langrock_flexible_2012} % 
--- used in animal ecology under the rubric of the ``multiphasic movement
framework'' \cite{fryxell_multiple_2008} --- that consider the effects
of organisms' \emph{internal} states; in particular, HMMs model animal
movement as though individual animals' movement behaviour 
at particular times is determined
by which of a discrete set of unobserved movement states
(e.g. ``foraging'', ``traveling'', ``resting'') they currently occupy.
Conditional on the state occupied by an individual, HMMs typically
assume that animals follow a correlated random walk model
\cite{okubo_diffusion_1980,turchin1998quantitative}.

Ever-increasing capabilities of remote
sensors are making movement data available over an
ever-wider range of time scales, at both higher resolution (e.g. hourly data
from GPS collars vs. daily or weekly fixes for radio or VHF
collars) and longer extent (e.g. from a few days to months or years).  When analyzing such
long-term data, ecologists will more
often have to account for temporal variability in movement
behaviour at diurnal and seasonal scales that were previously
not captured in the data.

@@

Our second goal, therefore, is to explore whether
van de Kerk et al.'s results on optimal
model complexity might be driven at
least in part by structural problems with their statistical
model, i.e. the
assumption of temporally homogeneous behaviour.  For large data sets, 
information-theoretic model selection methods will
typically choose complex, highly parameterized models. When there is
only one way in which models can become more complex (e.g. by
increasing the number of latent states), complexity that is
present in the data but not accounted for in the model (e.g. spatial
or temporal heterogeneity) can be misidentified as other forms of
complexity.  We predict that increasing volumes of data will
increasingly lead researchers who are accustomed to fitting 
small models to sparse data into such traps.  We examine whether
allowing for diurnal variation in the Florida panther data allows us to
select models with fewer latent states; we also
fit models to simulated data with varying numbers of latent states,
and with and without temporal heterogeneity, to test our conjecture that
heterogeneity can be misidentified as behavioural complexity.

@@

van de Kerk et al. \cite{kerk2015hidden} considered time-homogeneous models with 
a variety of candidate distributions %
--- log-Normal, Gamma, and Weibull distributions for step lengths
and von Mises and wrapped Cauchy distributions for the turning angle --- %
concluding on the basis of the Akaike information criterion (AIC) 
that Weibull step length and wrapped Cauchy turning angle distributions
were best.  
Since our analysis aims for simplicity and qualitative
conclusions rather than for picking the very best
predictive model, we focus on models that treat each step as
a univariate, log-Normally distributed observation, glossing
over both the differences in shape between
the three candidate step-length distributions and the effects
of considering multivariate (i.e., step length plus turning angle)
observations.  To check that this simplification
does not distort our conclusions
we do briefly compare log-Normal and Weibull
step-length distributions, with and without 
a von Mises-distributed turning angle included in the model (Figure 2).
(Note that most movement analyses, including van de Kerk et al.
\cite{kerk2015hidden}, are only partially multivariate, 
treating step length and turning angle at a particular time
as multivariate observations for the purpose of HMM
analysis but neglecting possible correlations
between the two measures.)

@@

Using BIC also simplifies 
evaluation of model selection procedures; 
it is easier to test whether our model selection
procedure has selected the model used to simulate the data,
rather than testing whether it has selected the model with
the minimal Kullback-Leibler distance
\cite{Richards2005}. We recognize that 
ecologists will often be interested in maximizing predictive
accuracy rather than selecting a true model, and that as
usual in ecological systems the true model will be far more
complicated than any candidate model \cite{BurnhamAnderson1998}.
We have repeated some of our analyses using AIC rather than BIC
(not shown); for our examples, the qualitative conclusions stated here
for BIC-optimality carry over to analyses using AIC.

@@

We ran a simulation experiment in which we fitted HMMs with
both homogeneous and heterogeneous 
transition probabilities to simulated data with
heterogeneous transition probabilities, to see whether
the correct (heterogeneous-transition) model correctly
identified the number of states while the misspecified
(homogeneous-transition) model overestimated the number of states.
We used 100 realizations of a two-state HMM with sinusoidally varying
transition probabilities and fitted it with HMMs ranging from
2 to 5 hidden states, with and
without temporal heterogeneity in the transition probabilities.

@@

Model complexity and the number of parameters increase as the number of latent states increase.
For a fixed number of states homogeneous FMMs are simplest, followed by homogeneous HMMs and finally by FMMs and HMMs incorporating temporal heterogeneity. In general, the number of free parameters in an HMM is the sum of the number of free parameters for each of the three model components. Let $n$ be the number of hidden states and $k_{i}, k_{t}, k_{e}$ be the number of parameters describing the covariate-dependence of the prior distribution, transition function and emission distributions; that is, for a homogeneous model, $k=1$, while a single numeric covariate or a categorical predictor with two levels would give $k=2$. Then the number of free parameters of an HMM is:
$k_{i}\cdot (n-1)$ (\emph{Initial states})
$+ k_{t}\cdot n\cdot (n-1)$ (\emph{Transition probabilities})
$+ k_{e}\cdot n$ (\emph{Emission parameters}).
As the number of states increases, the number of free parameters in (homogeneous or heterogeneous) FMMs and time-homogeneous HMMs increases linearly, whereas for HMMs with temporal heterogeneity (or covariate-dependent transitions more generally) the number increases quadratically. When comparing BICs, it is important to account for the tradeoff between log-likelihood and number of states, but also log-likelihood and number of free parameters.

We used simulations to predict expected hourly step lengths and
autocorrelation functions (ACF).  While the computation of expected
step length and ACF is straightforward for FMMs, and feasible for
homogeneous HMMs, the interaction between the geometric dwell time
within each state and the temporally varying interaction probabilities
makes it infeasible for more complex models.  We used this approach to
validate our models, comparing our simulated predictions with the
observed movements.  The more usual approach, generating predictions
from the expected step lengths conditional on the most likely state
sequence predicted by the Viterbi algorithm or pseudo-residuals \cite{zucchini_hidden_2009,langrock_flexible_2012}, is somewhat problematic
because the predictions by these methods rely on the observed data. 
This approach is useful to predict
missing data in the observation sequence, but because it is
conditional on the observed values, it can not reliably evaluate
goodness of fit for HMM models with
different degrees of structural complexity.

@@

%% fig 1 goes somewhere here
The simulation supports our hypothesis that homogeneous transition HMMs can overestimate the number of hidden states. Heterogeneous transition models can always predict the correct number of states, whereas the temporally homogeneous models overestimate the number of states (based on BIC-optimality: Figure 1).

\bmb{need to describe in words what the figure says}

@@

Finally, if we want to estimate the number of states, BIC is not necessarily 
good for estimation of number of states \cite{biernacki2000assessing}, 
but it can be useful as an approximate upper limit estimate. This is a common problem when using BICs, thus, the "knee point" (jump point) was proposed for BIC in partitioning based clustering via angle-based method \cite{zhao_knee_2008}. A similar method was used with likelihood \cite{dean2012behavioural} where both are applicable in our case, and suggest fewer number of states, but we do not know the reliability when fitting more complex models (IA. the shape of the BIC curve can be affected by the non-linear increase of parameters with every additional state).

Incorporating temporal heterogeneity in animal movement is one step in
the right direction, but much remains to be done. 
Our model neglects other predictors, such as sex, habitat type or
location with respect to environmental features such as roads,
that can potentially improve goodness of fit and predictions and 
further reduce the estimated number of states. While adding
more covariates is in principle straightforward using existing
frameworks, including all possible biological complexities in 
a HMM with state-dependent transitions may rapidly become intractable
in terms of both computational time and complexity of choosing
among possible reduced models and numbers of states.
Better diagnostic procedures and tests are needed:
these can both test overall goodness-of-fit \cite{potts_generalized_2014} 
and, more importantly, localize fitting problems to particular aspects of the
data so that models can be constructed without needing to
include all possible features of interest. Because there are a huge number of potential complexities that can be added to movement models (e.g. spatial/temporal/among-individual heterogeneity; effects of conspecific attraction or avoidance; memory or cognitive effects), each with associated costs in researcher and computational effort, such diagnostic plots are invaluable.

@@

\clearpage
